Parsing Postfix log files

### 2009-01-11 doc/logparser-other-parsers-review.tex
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Cleanups and improvements to the review of other parsers.

\section{State of the Art Review}

@@

It is important to compare and contrast newly developed programs,
algorithms, and parsers against those already available, to accurately
judge what improvements, if any, are delivered by the newcomers.  Ten other
parsers have been reviewed as part of the background research for this
project, starting with the only previously published research in this area
that the author is aware of.  There are not that many previously developed
Postfix log file parsers --- indeed it was quite difficult to find ten
parsers to review for this project --- and the functionality offered ranges
from quite basic to much more mature, depending on the needs of the author
of the parser.  The programs are not reviewed from an independent
viewpoint; the objective is to compare and contrast each program with
\parsername{}.  There are some important differences between \parsername{}
and the parsers reviewed here:

@@

Each of the reviewed parsers was tested with the \numberOFlogFILES{} test log files
described in \sectionref{parser efficiency}.  The data extracted by
\parsername{} is documented in \sectionref{connections table} and
\sectionref{results table}; that list is repeated here, without the
explanations:  server\_ip, server\_hostname, client\_ip, client\_hostname,
helo, queueid, start, end, warning, smtp\_code, sender, recipient, size,
message\_id, data, timestamp.

\subsection{Log Mail Analyser}

@@

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
main table the key is an integer referred to by secondary tables, and the
value is a \gls{CSV} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
sender/recipient/date/\gls{IP} as the key, and the value is a \gls{CSV}
list of integers referring to the main table.  This effectively
re-implements \gls{SQL} foreign keys, but without the functionality offered
by even the most basic of \gls{SQL} databases (joins, ordering, searches,
etc.).  It also requires custom code to search on some combination of the
above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the
least useful of the three output formats: it does not provide the
functionality of a basic \gls{SQL} database, not can it be used with
standard Unix text processing tools.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an \gls{SQL} database, unlike with
Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
and provide commands for displaying it.  Berkeley DB does not embed a
schema, as there is neither requirement nor benefit; it only provides
\textbf{(key, value)} pairs, so any additional structuring of the data is
imposed by the application, thus it behoves the application to document
this structure.

@@

\gls{LMA} stores the following date: time and date, client hostname and
\gls{IP} address, server \gls{IP} address, sender and recipient addresses,
\gls{SMTP} code, and size (for accepted mails only).  It omits XXX XXX\@.
Handling of multiple recipients, \gls{SMTP} codes or remote
servers\footnote{A single mail may be sent to multiple remote servers if it
was addressed to recipients in different domains, or Postfix needs to try
multiple servers for one or more recipients.} is not explained in the
paper; experimental observation shows that multiple records are added when
there are multiple recipients (sadly the records are not associated or
linked in any way), and presumably the same approach is taken when there
are multiple destination servers.

@@

\subsection{Pflogsumm}

@@

Pflogsumm produces a report designed for troubleshooting rather than for
in-depth analysis.  It does not support saving any data or extracting more
data than is required for producing the report.  Both the parsing and
reporting are difficult to extend as it is a specialised tool, unlike the
easily extensible design of \parsername{}.  It does not correlate log lines
by queueid or \gls{pid}, and does not need to deal with the complications
encountered during this project.  Pflogsumm produces a useful report, and
successfully parsed the \numberOFlogFILES{} log files tested
with,\footnote{The results it reported were not verified in detail, but it
did not report any errors, and has a very good reputation amongst Postfix
users.} but it is not a suitable base for this project.

Pflogsumm's report includes the following:

@@

Many options are available to include or exclude certain sections of the
report.

\subsection{Sawmill Universal Log File Analysis and Reporting}

@@

Sawmill is a general purpose commercial product that parses 818 log file
formats (as of 2008/11/23) and produces reports from the extracted data.
Its data extraction facilities (described later) are too limited to save
sufficient data for the purposes of this project; it extracts three
different sets of data from Postfix log files, but they are not interlinked
in any way.  The documentation does not suggest that any attempt is made to
correlate log lines or deal with the difficulties documented in
\sectionref{complications}.

Sawmill has three different Postfix log file parsers:

@@

Even if the three data sets were linked together Sawmill would extract less
data than \parsername{}, and it does not appear to extract data about
rejections except when the rejection is caused by a \gls{DNSBL} check
(\textit{RBL list\/} in the list of fields).

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema.  A
commercial version of MySQL is required due to MySQL licensing
restrictions, but Sawmill's documentation explains why and includes
instructions on how to compile Sawmill so that it can use a non-commercial
version of MySQL\@.

Sawmill's web interface supports searching on any combination of the fields
it extracts, but the interface is neither as simple to use nor as
informative as the interface provided by Splunk.  The administrative
interface is much easier to use than Splunk's: it took only five minutes to
start parsing a directory of log files.  Searches performed using the web
interface produced reasonable results, which were not verified in detail.

@@

XXX ADD A SUMMARY\@.

\subsection{Splunk}

@@

Splunk aims to index all an organisation's log files, providing a
centralised view capable of searching and correlating diverse log sources.
The web interface supports complicated searches, providing statistics and
graphs in real time, a facility not provided by \parsername{} (report
generation has been deferred to a subsequent program).  Saved searches can
be run periodically and the results emailed to a recipient or sent to an
external program for further processing (though maybe without the graphs
and detailed statistics); the author was unable to save searches, though
that may have been due to limitations in the free version.  The database is
not available for use by external programs, whereas \parsername{} provides
the database and leaves it to the user to utilise it without limit or
restriction.  The interface is optimised for interactive rather than
automated queries and it does not appear to be possible to write
independent tools to utilise the Splunk database; there are additional
reports available on \url{http://www.splunkbase.com/}, but the author was
unable to find any: every category was empty, even those that the interface
claimed had reports available.  Many types of reports are bundled with the
software, though most are variations of a bar or pie chart, except bubble
and heatmap graphs.  It is easy to drill down through the graphs to extract
a portion of the data (e.g.\ select the hour with the largest number of
events, then select a particular host, and finally a specific address),
though it is not possible to search on partial words.  All searches
performed using the indexed data returned reasonable results.  The web
interface is quite attractive and simple to use when searching, but as an
administrator it seems unnecessarily difficult to perform simple tasks.

When testing it took roughly 30 minutes to add a single log file to be
indexed (log files must be indexed before searching), with the downside
that the log file was copied into a spool directory, doubling the disk
space usage.  The most suitable test would be to index all the log files in
a particular directory, but after three hours, numerous futile attempts,
and reading all the available documentation the author admitted defeat.
Using the \gls{CLI} was more successful: invoking the command \newline{}
\tab{} \texttt{splunk find logs }\textit{log-directory\/}\newline{} added
40 of the \numberOFlogFILES{} log files to the queue for indexing.
Repeated attempts enqueued the same 40 log files, without explaining why
the others were excluded.\footnote{The log files appear to have been
indexed once only; presumably Splunk keeps track of the files it has
indexed and discards requests to index files for a second time.  This may
or may not be a useful feature for \parsername{}.} There did not appear to
be an option to order the log files to ensure they would be processed in
the order they were created, though this may not be necessary or beneficial
with Splunk.  Subsequently the author was successful in adding a single
file at a time using the \gls{CLI}; a simple loop to add all desired log
files was sufficient to index all files.  Splunk will periodically check
all those files for updates unless they are manually removed from its list;
this may or may not be useful behaviour.  Splunk did not appear to have any
difficulty in parsing the log files, once instructed to do so.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

Copious documentation is made available on \url{http://www.splunk.com/},
but poor organisation and sheer abundance makes it extremely hard to find
useful information.  Searching using the website's interface confusingly
tends to return results from old documentation rather than new.  In
general the documentation appears to have been written by someone
intimately acquainted with the software and having difficulty understanding
how a newcomer would approach tasks or the questions they would ask.

@@

When parsing Postfix log files Splunk extracts some standard fields: to and
from addresses, HELO hostname, time and date, the host the log files were
generated on, and protocol (\gls{SMTP} or \gls{ESMTP}).  It parses the
standard syslog\glsadd{syslog} fields at the beginning of the log line, and
extracts any \texttt{key=<value>} pairs occurring after the standard syslog
prologue; these pairs are the fields listed above.  Searches can be based
on the extracted fields or the full text of the log line.  \parsername{}
extracts noticeably more data, though it does not make the full text of the
line available, and the full power of \gls{SQL} is available when
searching, allowing the user to search on arbitrarily complicated
conditions.

@@

\subsection{Isoqlog}

@@

\noindent{}\url{http://www.enderunix.org/isoqlog/} \newline{}
(Last checked 2007/08/13.)

@@

XXX FINISH THIS\@.

\subsection{AWStats}

@@

\noindent{}\url{http://awstats.sourceforge.net/} \newline{}
\url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
(Last checked 2007/10/13.)

AWStats will produce simple graphs for many different services, but
supporting many different services without special purpose code limits its
functionality.  The data it will extract from an \gls{MTA} log file is
limited in comparison to \parsername{}: \newline{} \tab{} time2, email,
email\_r, host, host\_r, method, url, code, and bytesd.\newline{} XXX WHAT
FIELDS ARE MISSING\@?  There does not seem to be an explanation of any of
those fields in the documentation (\parsername{} provides copious
documentation).  AWStats coerces Postfix log files into Apache\footnote{The
Apache web server is the most popular HTTP server in use over the past 10
years; more information is available at \url{http://httpd.apache.org/}.}
format log files, for analysis by AWStats' HTTP log file parser.  The
converting parser only deals with a small portion of the log lines
generated by Postfix, silently skipping those it cannot deal with, and does
not distinguish between different types of rejection; it would be a lot of
work to extend it to handle new log lines.  Although it does correlate log
lines by queueid (not by pid), it does not deal with any of the other
complications described in \sectionref{complications}.  AWStats supports
saving data but the format of the saved data is not documented, as far as
the author could tell.  It also supports reading compressed log files, but
that functionality was not tested.

@@

XXX NEED A CONCLUSION\@.

\subsection{Anteater}

@@

\subsection{Yet Another Advanced Logfile Analyser}

@@

\subsection{Lire}

@@

\subsection{Logrep}

@@

\subsection{Summary}

>>>>

\chapter{State of the Art Review}

@@

It is important to compare and contrast newly developed programs,
algorithms, and parsers against those already available, to accurately
judge what improvements, if any, are delivered by the newcomers.  Ten
parsers have been reviewed as part of the background research for this
project, comparing and contrasting each parser with \parsername{}; the
first parser reviewed is the only previously published research in this
area that the author is aware of.    There are not many previously
developed Postfix log file parsers --- indeed it was quite difficult to
find ten parsers to review for this project --- and the functionality
offered ranges from quite basic to much more mature, depending on the needs
of the author of the parser.  There are some important differences between
\parsername{} and the parsers reviewed here:

XXX SHOULD THIS BE IN THE CONCLUSION INSTEAD\@?

@@

Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
log files described in \sectionref{parser efficiency}.  The data extracted
by \parsername{} is documented in \sectionref{connections table} and
\sectionref{results table}; for convenience that list is repeated here:
server\_ip, server\_hostname, client\_ip, client\_hostname, helo hostname,
queueid, start time, end time, smtp\_code, sender, recipient, size,
message\_id.

\section{Log Mail Analyser}

@@

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
main table the key is an integer referred to by secondary tables, and the
value is a \gls{CSV} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
sender/recipient/date/\gls{IP} as the key, and the value is a \gls{CSV}
list of integers referring to the main table.  This effectively
re-implements \gls{SQL} foreign keys, but without the functionality offered
by even the most basic of \gls{SQL} databases (joins, ordering, searches,
etc.).  It also requires custom code to search on some combination of the
above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the
least useful of the three output formats: it does not provide the
functionality of a basic \gls{SQL} database, nor can it be used with
standard Unix text processing tools.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an \gls{SQL} database, unlike with
Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
and provide commands for displaying it.  Berkeley DB does not embed a
schema, as there is neither requirement nor benefit; it only provides
\textbf{(key, value)} pairs, so any additional structuring of the data is
imposed by the application, thus the application must document this
structure.

@@

\gls{LMA} stores the following date: time and date, client hostname and
\gls{IP} address, server \gls{IP} address, sender and recipient addresses,
\gls{SMTP} code, and size (for accepted mails only); unlike \parsername{}
it does not store the server hostname, helo hostname, queueid, start and
end times, or message id (for accepted mails only).  Handling of multiple
recipients, \gls{SMTP} codes or remote servers\footnote{A single mail may
be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper; experimental observation shows
that multiple records are added when there are multiple recipients (sadly
the records are not associated or linked in any way), and presumably the
same approach is taken when there are multiple destination servers.

@@

\section{Pflogsumm}

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data than is not required for producing its report.  Both the
parsing and reporting are difficult to extend as it is a specialised tool,
unlike the easily extensible design of \parsername{}.  It does not
correlate log lines by queueid or \gls{pid}, and does not need to deal with
the complications encountered during this project.  Pflogsumm produces a
useful report, and successfully parsed the \numberOFlogFILES{} log files
tested with,\footnote{The results it reported were not verified in detail,
but it did not report any errors, and has a very good reputation amongst
Postfix users.} but it is not a suitable base for this project.  It does
not extract the helo hostname, queueid, start and end times, or message id.

Pflogsumm has many options are available to include or exclude certain
sections of the report; by default it includes the following:

@@

\section{Sawmill Universal Log File Analysis and Reporting}

@@

Sawmill is a general purpose commercial product that parses 818 log file
formats (as of 2008/11/23) and produces reports from the extracted data.
Its data extraction facilities (described later) are too limited to save
sufficient data for the purposes of this project: although it can extract
three different sets of data from Postfix log files, they are not
interlinked in any way.  The documentation does not suggest that any
attempt is made to correlate log lines or deal with the difficulties
documented in \sectionref{complications}.

Sawmill has three different Postfix log file parsers, extracting three
different sets to data:

@@

Even if the three data sets were combined Sawmill would extract less data
than \parsername{}: it omits the helo hostname, queueid, and start and end
times.  Sawmill does not extract any data about rejections except when the
rejection is caused by a \gls{DNSBL} check (\texttt{RBL list} in the list
of fields).

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema.  A
commercial version of MySQL is required due to MySQL licensing
restrictions, but Sawmill's documentation explains why and includes
instructions on how to compile Sawmill so that it can use a non-commercial
version of MySQL (this was not attempted during the review process).

Sawmill's web interface supports searching on any combination of the fields
it extracts, and the results obtained from searches were accurate.  The
interface for searching is neither as simple to use nor as informative as
the interface provided by Splunk.  The administrative interface is much
easier to use than Splunk's: it took only five minutes to start parsing a
whole directory of log files.  Searches performed using the web interface
produced reasonable results, which were not verified in detail.

@@

In summary Sawmill suffers from being a general purpose product; it is
probably much more useful when parsing log files where each log line is
self-contained, rather than linked to other log lines (e.g.\ web server log
files).  It is not suitable as a base for this parser, as the source code
made available is obfuscated and not intended for modification; in addition
the architecture would probably need to be overhauled to deal with
correlating log lines.

\section{Splunk}

@@

Splunk aims to index all of an organisation's log files, providing a
centralised view capable of searching and correlating diverse log sources.
The web interface supports complicated searches, providing statistics and
graphs in real time, a facility not provided by \parsername{} (report
generation has been deferred to a subsequent program).  Saved searches can
be run periodically and the results emailed to a recipient or sent to an
external program for further processing (though maybe without the graphs
and detailed statistics); the author was unable to save searches, though
that may have been due to limitations in the free version.  The database is
not available for use by external programs, whereas \parsername{} provides
the database and leaves it to the user to utilise it without limit or
restriction.  The interface is optimised for interactive use rather than
automated queries and it does not appear to be possible to write
independent tools to utilise the Splunk database; there are additional
reports available on \url{http://www.splunkbase.com/}, but the author was
unable to find any: every category was empty, even those that the interface
claimed had reports available.  Many types of reports are bundled with the
software, though most are variations of a bar or pie chart, except bubble
and heatmap graphs.  It is easy to drill down through the graphs to extract
a portion of the data (e.g.\ select the hour with the largest number of
events, then select a particular host, and finally a specific address),
though it is not possible to search on partial words.  All searches
performed using the indexed data returned reasonable results.  The web
interface is quite attractive and simple to use when searching, but as an
administrator it seems unnecessarily difficult to perform simple tasks.

When testing Splunk it took roughly 30 minutes to figure out how to add a
single log file to be indexed so that it could be searched, with the
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, numerous
futile attempts, and reading all the available documentation, the author
admitted defeat.  Using the \gls{CLI} rather than the web interface was
more successful: invoking the command \newline{} \tab{} \texttt{splunk find
logs }\textit{log-directory\/}\newline{} added 40 of the
\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
enqueued the same 40 log files, without explaining why the others were
excluded.\footnote{The log files appear to have been indexed once only;
presumably Splunk keeps track of the files it has indexed and discards
requests to index files for a second time.  This may or may not be a useful
feature for \parsername{}.} There did not appear to be an option to ensure
the log files would be processed in the order they were created, though
this may be neither necessary nor beneficial with Splunk.  Subsequently the
author was successful in adding a single file at a time using the
\gls{CLI}; a simple loop to add all desired log files was sufficient to
index all files.  Splunk will periodically check all those files for
updates unless they are manually removed from its list; this may or may not
be useful behaviour.  Splunk did not appear to have any difficulty in
indexing the log files, once they had been successfully added to its queue.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

Copious documentation is made available on \url{http://www.splunk.com/},
but poor organisation and sheer abundance makes it extremely hard to find
useful information.  Searching using the website's interface confusingly
tends to return results from old documentation rather than new.  In general
the documentation appears to have been written by someone intimately
acquainted with the software who has difficulty understanding how a
newcomer would approach tasks or the questions they would ask.

@@

When parsing Postfix log files Splunk extracts some standard fields: to and
from addresses, HELO hostname, time and date, the host the log files were
generated on, and protocol (\gls{SMTP} or \gls{ESMTP}).  It parses the
standard syslog\glsadd{syslog} fields at the beginning of the log line, and
extracts any \texttt{key=<value>} pairs occurring after the standard syslog
prologue; these pairs are the fields listed above.  Searches can be based
on the extracted fields or the full text of the log line.  \parsername{}
extracts noticeably more data (client and server \gls{IP} and hostname,
queueid, start and end times, \gls{SMTP} code, and message ID), though it
does not make the full text of the line available (this could be trivially
added if desired, but would greatly increase the size of the resulting
database).  The full power of \gls{SQL} is available when searching the
data extracted by \parsername{}, allowing the user to search on arbitrarily
complicated conditions.

@@

\section{Isoqlog}

@@

\noindent{}\url{http://www.enderunix.org/isoqlog/} \newline{}
(Last checked 2009/01/11.)

@@

\section{AWStats}

@@

\noindent{}\url{http://awstats.sourceforge.net/} \newline{}
\url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
(Last checked 2009/01/11.)

AWStats will produce simple graphs for many different services, but
supporting many different services without special purpose code limits its
functionality.  The data it will extract from an \gls{MTA} log file is
limited in comparison to \parsername{}: time2, email, email\_r, host,
host\_r, method, url, code, and bytesd.  There is no explanation for any of
those fields in the documentation (\parsername{} provides copious
documentation), so the author could not understand the extracted data, nor
determine what data is missing in comparison to \parsername{}.  AWStats
coerces Postfix log files into Apache\footnote{The Apache web server is the
most popular HTTP server in use over the past 10 years; more information is
available at \url{http://httpd.apache.org/}.} format log files, for
analysis by AWStats' HTTP log file parser.  The converting parser only
deals with a small portion of the log lines generated by Postfix, silently
skipping those it cannot deal with, and does not distinguish between
different types of rejection; it would be a lot of work to extend it to
handle new log lines.  Although it does correlate log lines by queueid (not
by pid), it does not deal with any of the other complications described in
\sectionref{complications}.  AWStats supports saving data but the format of
the saved data is not documented, as far as the author could tell.  It also
supports reading compressed log files, but that functionality was not
tested.

@@

AWStats was not a suitable base for this project, because it assumes that
all log files can be rewritten to be compatible with web server log files,
and will contain similar data; coercing Postfix log files into web server
log files, without substantial data loss, would require fully parsing the
Postfix log files, so AWStats would not be required.  It may be possible to
use AWStats graphing capabilities to generate reports, by generating log
files to use as input to AWStats from the data extracted by \parsername{}.

\section{Anteater}

@@

\section{Yet Another Advanced Logfile Analyser}

@@

\section{Lire}

@@

\section{Logrep}

@@

\section{Summary}
